---
title: "R Notebook"
output: html_notebook
---

In this version, I use the new function that was used to compare Josh's and my results. 

This fix may work for all years. At least, 2015 and 2016 files were processed without any errors. Results need to be compared with Josh's outputs. The old extraction missed many things, although the differences are minor. I think it's now (December 2023) a lot better extraction method than old versions. I tested it for data since 2010 (2010, 2011, 2015, 2016, 2020, 2022, 2023). They worked fine and picked up some strange data, which were fixed in the input files. 

The recent vesion as of December 2023 also creates input files for Laake's approach. That approach requires double-observer data, which are only available for 2010 and 2011 in the recent data. Years without double observers are adjusted by using the average observation probability from years with double observers. Use LaakeAnalysis_NewData.R for the analysis. It is incomplete as of December 2023. 


The station information needs to be added. The data files that I have don't have that info... this needs to be fixed in the future. This has been fixed. 2023-12-12. Not exactly... They are not entered in the n array correctly. Unfortunately, the current WinBUGS approach does not let effort duration between station 1 and 2 (the effort array is 2D, whereas n array is 3D). If we were to use these data, the effort array needs to be changed to 3D and their timing match with the station 1 - the assumed N is the same for station 1 and 2. This is not an easy fix. I'm not sure how Durban took care of this issue for 2009/2010 and 2010/2011 data. 


```{r}

rm(list=ls())
library(tidyverse)
library(lubridate)
source("Granite_Canyon_Counts_fcns.R")

save.output <- F #TRUE
# https://statisticsglobe.com/dplyr-message-summarise-has-grouped-output-r
options(dplyr.summarise.inform = FALSE)

# I forget why I have .1 convention in the subdirectory name. But, I decided to keep it.
#out.dir <- "RData/V2.1_Nov2024"
out.dir <- "RData/V2.1_Feb2025"
if (!dir.exists(out.dir)) dir.create(out.dir)

# YEAR should be one of 2010, 2011, 2015, 2016, 2020, 2022, 2023 (and beyond)
# They are the end year of the season, which is the season name. For example, the 
# 2020/2021 season is called the 2021 season.

YEAR <- 2024 #10 #Enter the year of the data files

# For 2009/2010 (maybe in the previous years too but I don't have any of those files),
# Time is entered in decimal hours, not hh:mm:ss. So, that needs to be accommodated in
# data extraction code. 2022-10-05

shift_dur_min <- 90   # 90 minutes
grace_min <- 80#5 # 60 #       # this makes it minimum of shift_dur_min - grace_min

# For Mar 2023 analysis, I changed this to minimum duration of 30 minutes
max_dur <- shift_dur_min + 5
min_dur <- shift_dur_min - grace_min
#min_dur <- 30

```


```{r}

if (YEAR == 2023){
  FILES <- list.files(path = paste0("Data/", YEAR, "/"),
                      pattern = "EditedGW")
  
} else {
  # if FILES input was not provided, make a list of files in the directory. This should work
  # for other years 
  FILES <- list.files(paste0("Data", "/", YEAR, "/"))

}
  
all.effort.list.1 <- Data_Out.list.1 <- all.sightings.list.1 <- list()
all.effort.list.2 <- Data_Out.list.2 <- all.sightings.list.2 <- list()

# The next for loop throws back "NAs introduced by coercion" but not to worry.
# These are results of as.numeric() for the line numbers, where some lines may
# be non-numbers, e.g., "Edited" and other text. Also, some BF and VS codes are
# missing and they return warnings "in 'transmute()'..." They turn into NAs so 
# they are not problems. But... it'd be best to remove the first comment line.
# I fixed the first line issue in Dec 2023.

# 2023-11-28: Shift IDs are incorrect. They need to be defined from 7:30, rather than
# the first shift of each day. Need to... extract all data first using get.data.
# combine them to make one data frame, then send that to get.shift by selecting 
# one day at a time. Need to change the ff loop and create another loop (or apply)
# to extract shifts/day. Fixed. Now has a variable named Shift with upper S, which
# defines shifts. 

data.list <- list()
ff <- 1
for(ff in 1:length(FILES)){ 
  #print(paste0("file = ", ff))
  data.list[[ff]] <- get.data("Data/", YEAR, FILES = FILES, ff = ff)
}

# combine all data
data.all <- do.call("rbind", data.list)
#data.all$Shift <- shift.definition(as.Date(data.all$V3, format = "%m/%d/%Y"), data.all$V4)

# The next loop doesn't work when there were two teams. 
# This needs to be fixed - 2023-12-04 
# This has been fixed
# 
# get.shift is not working for 2010 - Shift is not defined. 2023-12-05
# This has been fixed. I created a dataframe that defines the beginning
# and the end of each shift in Granite_Canyon_Counts_fcns.R 
# 
# 2023-12-08: Group ID 5 on Jan 1, 2010 was double counted in watch 4 and 5!
# This has been fixed. 

# 2023-12-11: There are still some issues for 2010 when shifts are not exactly
# 1.5 hr long and when observers change within a shift. get.shift function
# needs to be tested more thoroughly. I changed the part where effort and sightings
# are extracted for Laake's analysis (lines 691-928). That needs to be tested
# further. I think... this is all taken care of now. 2023-12-12

unique.dates <- unique(data.all$V3)
n.stations <- vector(mode = "numeric", length = length(unique.dates))
c <-2
for (c in 1:length(unique.dates)){
  #print(paste0("Date = ", c))
  data.all %>% 
    filter(V3 == unique.dates[c]) -> data  
  
  # If there were two stations, there should be two file names for that day
  n.stations[c] <- length(unique(data$ff))
  
  if (n.stations[c] == 2){
    data.1 <- filter(data, ff == unique(data$ff)[1])
    data.2 <- filter(data, ff == unique(data$ff)[2])
  } else {
    data.1 <- data    
  }
  
  Shifts.1 <- which(data.1$V2 %in% "P") #start of all shifts
  
  all.effort.1 <- Output.list.1 <- all.sightings.1 <- list()
  
  k <- 4
  for(k in 1:length(Shifts.1)){
    #print(paste0("Shift.p = ", k))
    out.shift.1 <- get.shift(YEAR, data.1, k)
    
    Output.list.1[[k]] <- out.shift.1$out.df
    all.sightings.1[[k]] <- out.shift.1$data  # added for Laake's analysis
    all.effort.1[[k]] <- out.shift.1$shift.effort  # for computing effort for Laake's analysis
  }
  
  Data_Out.list.1[[c]] <- do.call("rbind", Output.list.1)
  all.sightings.list.1[[c]] <- do.call("rbind", all.sightings.1) %>% filter(n > 0)
  all.effort.list.1[[c]] <- do.call("rbind", all.effort.1)
  
  if (n.stations[c] == 2){
    
    #data.2 <- filter(data, ff == unique(data$ff)[2])
    
    Shifts.2 <- which(data.2$V2 %in% "P") #start of all shifts
    all.effort.2 <- Output.list.2 <- all.sightings.2 <- list()
    k <- 6
  
    for(k in 1:length(Shifts.2)){
      #print(paste0("Shift.s = ", k))
      out.shift.2 <- get.shift(YEAR, data.2, k)
      
      Output.list.2[[k]] <- out.shift.2$out.df
      all.sightings.2[[k]] <- out.shift.2$data  # added for Laake's analysis
      all.effort.2[[k]] <- out.shift.2$shift.effort  # for computing effort for Laake's analysis
    }
    
    Data_Out.list.2[[c]] <- do.call("rbind", Output.list.2)
    all.sightings.list.2[[c]] <- do.call("rbind", all.sightings.2) %>% filter(n > 0)
    all.effort.list.2[[c]] <- do.call("rbind", all.effort.2)
        
  }
  
  #Sys.sleep(1)
}         

# For 2020, time was recorded in decimal hours, which need to be converted to H:M:S
# There is a function for that fractional_Hrs2HMS in Granite_Canyon_Counts_fcns.R

# For primary:
Data_Out_1 <- do.call("rbind", Data_Out.list.1) %>%
  mutate(station = "P")

if (YEAR == 2010){
  Data4Laake <- do.call("rbind", all.sightings.list.1) %>%
    mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
           Time_HMS = fractional_Hr2HMS(Time),
           Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift))
  
} else {
  Data4Laake <- do.call("rbind", all.sightings.list.1) %>%  
    mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
           Time_HMS = Time,
           Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift))
} 

Data4Laake  %>%
  transmute(Date = Date,
            Time_PST = Time_HMS,
            key = key.1,
            Group_ID = Group_ID,
            nwhales = n,
            beaufort = bft,
            vis = vis,
            Bearing = Bearing,
            Reticle = Reticle,
            Distance = Distance,
            shift = shift,
            date.shift = Date.shift,
            observer = toupper(Observer),
            begin = begin,
            end = end,
            time = time,
            effort = effort,
            station = "P") %>%
  filter(nwhales > 0) -> Data4Laake_1

# 2023-11-29 Double check on what the "key" should be in the effort file. In Laake's
# analysis, key is Date_Shift. 
Effort4Laake_1 <-  do.call("rbind", all.effort.list.1) %>%
  mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
         Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift)) %>%
  transmute(Date = Date,
            Start.year = YEAR - 1,
            key = key.1,
            shift = shift,
            date.shift = Date.shift,
            npods = npods,
            nwhales = nwhales,
            beaufort = bft,
            vis = vis,
            observer = Observer,
            begin = begin,
            end = end,
            time = time,
            effort = effort,
            pphr = npods/(24*effort),
            station = "P")


# Because I defined "shift" by observer changes, not by time definitions, there are 
# occasions where one pod was sighted over more than two shifts (e.g., 1/14/2010, 
# Group_ID = 66, secondary observer).
# These need to be dealt with by removing earlier recordings. This is done when 
# one or more pods are found in two consecutive shifts within get.shift function.
# 
Data4Laake_1 %>% 
  group_by(Date, Group_ID) %>%
  summarise(n = n()) %>%
  filter(n > 1) -> n_data.1

# Look at n_data.1, which lists groups that were recorded more than once. 
tmp.1 <- list()
k <- 1
if (nrow(n_data.1) > 0){
  for (k in 1:nrow(n_data.1)){
    Data4Laake_1 %>%
      filter(Date == n_data.1$Date[k],  
             Group_ID == n_data.1$Group_ID[k]) -> tmp.1[[k]]
    
  }
}

Effort4Laake_1$seq <- seq(from = 1, to = nrow(Effort4Laake_1)) 
Effort4Laake_1 %>%
  mutate(watch.key = paste0(Start.year, "_", seq)) %>%
  dplyr::select(-c(seq)) -> Effort4Laake_1

# pphr (pods per hour) is in the sightings data, rather than in the effort in 
# Laake's analysis. So, I move the variable.
# Warning: Detected an unexpected many-to-many relationship between `x` and `y`.
# Check this... 2024-02-23
Data4Laake_1 %>%
  left_join(Effort4Laake_1 %>%
              dplyr::select(key, pphr), by = "key") -> Data4Laake_1

# For secondary if exists:
Data_Out_2 <- do.call("rbind", Data_Out.list.2)
  
if (!is.null(Data_Out_2)){
  Data_Out_2 %>%
    mutate(station = "S") -> Data_Out_2
  
  if (YEAR == 2010){
    Data4Laake <- do.call("rbind", all.sightings.list.2) %>%
      mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
             Time_HMS = fractional_Hr2HMS(Time),
             Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift))
    
  } else {
    Data4Laake <- do.call("rbind", all.sightings.list.2) %>%  
      mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
             Time_HMS = Time,
             Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift))
  } 
  
  Data4Laake %>%
    transmute(Date = Date,
              Time_PST = Time_HMS,
              key = key.1,
              Group_ID = Group_ID,
              nwhales = n,
              beaufort = bft,
              vis = vis,
              Bearing = Bearing,
              Reticle = Reticle,
              Distance = Distance,
              shift = shift,
              date.shift = Date.shift,
              observer = toupper(Observer),
              begin = begin,
              end = end,
              time = time,
              effort = effort,
              station = "S") %>%
    filter(nwhales > 0) -> Data4Laake_2
  
  Data_Out <- rbind(Data_Out_1, Data_Out_2) %>%
    arrange(by = begin)
  
  Effort4Laake_2 <-  do.call("rbind", all.effort.list.2) %>%
    mutate(key.1 = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift, "_", key),
           Date.shift = paste0(as.Date(Date, format = "%m/%d/%Y"), "_", shift)) %>%
    transmute(Date = Date,
              Start.year = YEAR - 1,
              key = key.1,
              shift = shift,
              date.shift = Date.shift,
              npods = npods,
              nwhales = nwhales,
              beaufort = bft,
              vis = vis,
              observer = Observer,
              begin = begin,
              end = end,
              time = time,
              effort = effort,
              pphr = npods/(24*effort),
              station = "S")
  
  Effort4Laake_2$seq <- seq(from = 1, to = nrow(Effort4Laake_2)) 
  Effort4Laake_2 %>%
    mutate(watch.key = paste0(Start.year, "_", seq)) %>%
    select(-c(seq)) -> Effort4Laake_2
  
  # Because I defined "shift" by observer changes, not by time definitions, there are 
  # occasions where one pod was sighted over more than two shifts (e.g., 1/14/2010, 
  # Group_ID = 66, secondary observer).
  # These need to be dealt with by removing earlier recordings. This is done when 
  # one or more pods are found in two consecutive shifts within get.shift function.
  # 
  Data4Laake_2 %>% 
    group_by(Date, Group_ID) %>%
    summarise(n = n()) %>%
    filter(n > 1) -> n_data.2
  
  # Look at n_data.1, which lists groups that were recorded more than once. 
  tmp.2 <- list()
  k <- 1
  if (nrow(n_data.2) > 0){
    for (k in 1:nrow(n_data.2)){
      Data4Laake_2 %>%
        filter(Date == n_data.2$Date[k],  
               Group_ID == n_data.2$Group_ID[k]) -> tmp.2[[k]]
      
    }
  }
  
  # pphr (pods per hour) is in the sightings data, rather than in the effort in 
  # Laake's analysis. So, I move the variable.
  Data4Laake_2 %>%
    left_join(Effort4Laake_2 %>%
              select(key, pphr), by = "key") -> Data4Laake_2

  
  Data4Laake_all <- rbind(Data4Laake_1, Data4Laake_2) %>%
    arrange(by = begin)
  
  Effort4Laake_all <- rbind(Effort4Laake_1, Effort4Laake_2) %>%
    arrange(by = begin)  
} else {
  Data_Out <- Data_Out_1 %>%
    arrange(by = begin)
  
  Data4Laake_all <- Data4Laake_1 %>%
    arrange(by = begin)
  
  Effort4Laake_all <- Effort4Laake_1  %>%
    arrange(by = begin)  
}


```

2023-12-14 There were a few questionable entries for 2020. I emailed Aimée Lang on 2023-12-14 to verify the data. Email below:

"Please take a look at the data files attached (raw and edited). On Jan 6, 2020, Group 17 was first detected at 12:57:23. The same animal was detected at 15:27:54 (what?) and 15:33:00. Is it reasonable to change the group ID at 15:27:54, and the remaining groups for the day (18 and 19), accordingly?" 

She's on leave until 2023-12-19. 

2023-12-15 She agreed on my suggestion. Data file has been edited.

```{r}
# Check tmp.1 and tmp.2 for possible double counting of pods, when they were sighted 
# over more than two shifts. If one of these messages appear, check the corresponding
# list (i.e., tmp.1 or tmp.2). The raw data file may be edited. 
# 
if (length(tmp.1) > 0){
  print("Multiple sightings were found in primary observers")
  
}

if (exists("tmp.2")){
  if (length(tmp.2) > 0){
    print("Multiple sightings were found in secondary observers")
  }
  
}


```




```{r}
# Filter out short periods:

#Josh: 0.0625 is the exact watch period, but I've given some leeway. If it's less than 5 minutes short, I'm counting it
#If it's less than 10 minutes over the 1.5hrs, I'm also counting it (guessing that they forgot to log off or something)

#Entries that are less than 1.5hrs (5 minute grace period)
#shift_dur_min <- 90   # 90 minutes - 5 minutes
#grace_min <- 5
# Data_Out[which((Data_Out$end - Data_Out$begin) < (shift_dur_min - grace_min)/(24*60)),]
# 
# #Entries more than 1.5hrs (5 minute grace period)
# Data_Out[which((Data_Out$end - Data_Out$begin) > (shift_dur_min + grace_min)/(24*60)),] 
# 
# #Entries more than 1.5 hrs +/- 10 min
#grace_min <- 10
# Data_Out[which(Data_Out$end - Data_Out$begin < 
#                  (shift_dur_min - grace_min)/(24*60)),]
# Data_Out[which(Data_Out$end-Data_Out$begin > 
#                  (shift_dur_min + grace_min)/(24*60)),] 

# Final data
# Remove watches that were less than 85 minutes or greater than 95 minutes:
#grace_min <- 5

# For Mar 2023 analysis, I changed this to minimum duration of 30 minutes
#max_dur <- 90 + grace_min
#min_dur <- 90 - grace_min
#min_dur <- 30

Data_Out %>% 
  filter(dur > min_dur/(24*60) & 
           dur < max_dur/(24*60)) -> Correct_Length

Chaff <- Data_Out[which(Data_Out$end-Data_Out$begin < 
                          (shift_dur_min - grace_min)/(24*60)),]

Final_Data <- Correct_Length %>%
  filter(bf < 5, vs < 5)

WhalesDays <- Final_Data %>%
  filter(station == "P") %>%
  group_by(BeginDay) %>%
  mutate(PropDay = end-begin) %>%
  summarize(TotalWatch = sum(PropDay), TotalWhales=sum(n))

# ggplot(data = WhalesDays) + 
#   geom_point(aes(x = as.numeric(BeginDay), y = TotalWhales/TotalWatch))
#plot(x=WhalesDays$BeginDay,y=WhalesDays$TotalWhales/WhalesDays$TotalWatch)

ShiftsPerDay <- Final_Data %>%
  filter(station == "P") %>%
  group_by(BeginDay) %>%
  summarize(Watches = n())

```



```{r}

#Summary Stats for Report
Complete_Data <- Final_Data[complete.cases(Final_Data),]
Complete_Data$Eff <- Complete_Data$end - Complete_Data$begin

TotalHrs <- sum(Complete_Data %>% 
                  filter(station == "P") %>% 
                  dplyr::select(Eff))*24

TotalDays <- length(unique(floor(Complete_Data %>%
                                   filter(station == "P") %>%
                                   dplyr::select(begin))))

TotalObservers <- length(unique(Complete_Data$obs))

TotalWhales <- sum(Complete_Data %>%
                     filter(station == "P") %>%
                     dplyr::select(n))

WPH <- Complete_Data %>%
  filter(station == "P") %>%
  mutate(DaysSinceDec1 = as.numeric(BeginDay)) %>%
  group_by(DaysSinceDec1) %>%
  summarize(TotalWhales = sum(n), 
            TotalEffort = sum(Eff), 
            WPH = sum(n)/(sum(Eff)*24)) 

out.obj <- list(Data_Out = Data_Out,
                Correct_Length = Correct_Length,
                Final_Data = Final_Data,
                Complete_Data = Complete_Data,
                WPH = WPH,
                sightings_primary = all.sightings.list.1,
                sightings_secondary = all.sightings.list.2,
                Data4Laake = Data4Laake_all,
                Effort4Laake = Effort4Laake_all)

# Saving RDS file for Bayesian analyses (Durban and Eguchi)
# saveRDS(out.obj, file = paste0("RData/V2.1_Nov2023/out_", YEAR, "_Tomo_v2.rds"))
# saveRDS(out.obj, file = paste0("RData/V2.1_Nov2023/out_", YEAR, "_Tomo_v2.rds"))

if (save.output){
  #saveRDS(out.obj, file = paste0("RData/V2.1_Nov2023/out_", 
  #YEAR, "_min", min_dur, "_Tomo_v2.rds"))
  # saveRDS(out.obj, file = paste0("RData/V2.1_Feb2024/out_", 
  #YEAR, "_min", min_dur, "_Tomo_v2.rds"))
  
  saveRDS(out.obj, file = paste0(out.dir, "/out_", 
                                 YEAR, "_min", min_dur, 
                                 "_Tomo_v2.rds"))
  
  # Saving csv files for Laake analysis
  # write.csv(Data4Laake_all,
  #           file = paste0("Data/all_sightings_", YEAR, "_Tomo_v2.csv"))
  # write.csv(Effort4Laake_all,
  #           file = paste0("Data/all_effort_", YEAR, "_Tomo_v2.csv"))
  
}

```

