---
title: "Variance-covariance estimation"
output: html_notebook
date: "`r Sys.Date()`"
---

R notebook for extracting the variance-covariance matrix of abundance estimates.


```{r setup, include=FALSE}
rm(list = ls())

# source("Granite_Canyon_Counts_fcns.R")
library(tidyverse)
# library(lubridate)
library(flextable)
# library(readr)
# library(bayesplot)
# 

```


Calculate var-cov matrix

```{r Ver1-data, echo=FALSE, warning=FALSE, message= FALSE}
years <- "2007_2025"

seasons <- c("2006/2007", "2007/2008", "2009/2010", "2010/2011",
             "2014/2015", "2015/2016", "2019/2020", "2021/2022",
             "2022/2023", "2023/2024", "2024/2025")

x <- length(seasons)

# Bring in the WinBUGS output 
BUGS.output <- readRDS("RData/WinBUGS_2007to2025_v2_min85_100000_2025-02-25.rds")

# Extract posteriors for "Corrected.Est":
post.matrix <- BUGS.output$BUGS.out$sims.matrix
Corrected.Est.posterior <- post.matrix[, grep("Corrected.Est", dimnames(post.matrix)[[2]])]
var.cov <- var(Corrected.Est.posterior)
var.cov.log <- var(log(Corrected.Est.posterior))
dimnames(var.cov.log) <- list(seasons, seasons)

Summary.Stats <- BUGS.output$BUGS.out$summary
log.means.df <- data.frame(Season = seasons,
                           mean = as.vector(Summary.Stats[grep("Corrected.Est",
                                                               dimnames(Summary.Stats)[[1]]), "mean"]) %>%
                             log(),
                           median = as.vector(Summary.Stats[grep("Corrected.Est",
                                                                 dimnames(Summary.Stats)[[1]]), "50%"]) %>%
                             log())

means.df <- data.frame(Season = seasons,
                       mean = as.vector(Summary.Stats[grep("Corrected.Est",
                                                           dimnames(Summary.Stats)[[1]]), "mean"]),
                       median = as.vector(Summary.Stats[grep("Corrected.Est",
                                                             dimnames(Summary.Stats)[[1]]), "50%"]),
                       SD = as.vector(Summary.Stats[grep("Corrected.Est", 
                                                         dimnames(Summary.Stats)[[1]]), "sd"])) %>%
  mutate(CV = SD/mean)

```


Check to make sure CV for the most recent year matches what I provided in the report. 


```{r Table-nhats-v1, echo=FALSE, warning=FALSE}
flextable(means.df) %>% 
  set_caption(paste0("Mean, median, standard deviation (sd), and CV of marginal posterior distribution of corrected abundance of gray whales from the visual surveys off Granite Canyon, CA."))  %>%
  colformat_double(j = c("mean", "median"), digits = 0) %>%
  colformat_double(j = c("SD", "CV"), digits = 2) %>%
  set_table_properties(width = 0.5, layout = "autofit")
```


```{r}
# Function from Laake's code - log Normal theoretical confidence interval
conf.int = function(abundance, CV, alpha=0.05, digits=2, prt=FALSE){
  # Computes confidence intervals based on lognormal distr.
  # JMB / NMML / 11 Sep 2008
  
  if (alpha <0 || alpha > .999) stop("alpha must be in (0,1)")
  z = round(abs(qnorm(alpha/2)),2)
  if (prt) cat("N:",abundance,"  cv:",CV,"  alpha:",alpha,"  z:",z,"\n")
  C <- exp(z * sqrt(log(1 + CV^2)))
  SL <- round(abundance/C,digits)
  SU <- round(abundance * C,digits)
  data.frame(SL,SU)
}

logN.CI.95 <- conf.int(abundance = means.df$mean,
                       CV = means.df$CV,
                       alpha = 0.05) %>%
  transmute(Season = seasons,
            Mean = means.df$mean,
            Lower = SL,
            Upper = SU)



```

In the report, I don't provide SD or CV. In the following table, I compute the theoretical 95% confidence intervals using the log normal distribution. These confidence limits should be similar to the posterior 95% credible intervals in the report. 


```{r Table-95CI1, echo=FALSE, warning=FALSE}
flextable(logN.CI.95) %>% 
  set_caption(paste0("Upper and lower limits of theoretical confidence limits using the log-normal distribution."))  %>%
  colformat_double(j = c("Lower", "Upper"), digits = 1) %>%
  set_table_properties(width = 0.5, layout = "autofit")
```


These look good. So, create a text file with var-cov. Then, send it to Andre P.

```{r}

out.file.cov <- paste0("Data/var_cov_", years, ".dat")
out.file.mean.log <- paste0("Data/means_log_", years, ".dat")
out.file.mean <- paste0("Data/means_", years, ".dat")

write.table(var.cov.log, file = out.file.cov, quote = FALSE)

write.table(log.means.df, file = out.file.mean.log, quote = FALSE)

write.table(means.df %>%
              select(-c("SD", "CV")), file = out.file.mean, quote = FALSE)

```


